---
title: "R Notebook"
---




```{r}
library(SingleCellExperiment)
library(tidyverse)
```


```{r}
set.seed(1)
```


Data was downloaded from https://singlecell.broadinstitute.org/single_cell/study/SCP1663

Unfortunately, this cannot be automated, so you will manually need to download the files

* predictive_variable_slideseq_j20_rep1.rds
* predictive_variable_slideseq_j20_rep2.rds
* predictive_variable_slideseq_j20_rep3.rds
* predictive_variable_slideseq_j20_rep4.rds
* SpatialRNA_cropped_slideseq_j20_rep1.zip
* SpatialRNA_cropped_slideseq_j20_rep2.zip
* SpatialRNA_cropped_slideseq_j20_rep3.zip
* SpatialRNA_cropped_slideseq_j20_rep4.zip

Unzip the zip files. Those will contain RDS files (without the file extension). Lastly, move the files into the `data/cable_spatial_plaque_data` folder.

```{r}
rep1 <- readRDS("../data/cable_spatial_plaque_data/SpatialRNA_cropped_slideseq_j20_rep1")
rep2 <- readRDS("../data/cable_spatial_plaque_data/SpatialRNA_cropped_slideseq_j20_rep2")
rep3 <- readRDS("../data/cable_spatial_plaque_data/SpatialRNA_cropped_slideseq_j20_rep3")
rep4 <- readRDS("../data/cable_spatial_plaque_data/SpatialRNA_cropped_slideseq_j20_rep4")

plaque_dens_rep1 <- readRDS("../data/cable_spatial_plaque_data/predictive_variable_slideseq_j20_rep1.rds")
plaque_dens_rep2 <- readRDS("../data/cable_spatial_plaque_data/predictive_variable_slideseq_j20_rep2.rds")
plaque_dens_rep3 <- readRDS("../data/cable_spatial_plaque_data/predictive_variable_slideseq_j20_rep3.rds")
plaque_dens_rep4 <- readRDS("../data/cable_spatial_plaque_data/predictive_variable_slideseq_j20_rep4.rds")
```




```{r, paged.print = FALSE}
common_gene_names <- intersect(intersect(rownames(rep1@counts), rownames(rep2@counts)), intersect(rownames(rep3@counts), rownames(rep4@counts)))

merged_mat <- cbind(
  rep1@counts[common_gene_names, ], 
  rep2@counts[common_gene_names, ], 
  rep3@counts[common_gene_names, ], 
  rep4@counts[common_gene_names, ]
)
colnames(merged_mat) <- c(paste0(colnames(rep1@counts), "-rep1"),
                          paste0(colnames(rep2@counts), "-rep2"),
                          paste0(colnames(rep3@counts), "-rep3"),
                          paste0(colnames(rep4@counts), "-rep4"))

sce <- SingleCellExperiment(list(counts = merged_mat), colData = bind_rows(rep1@coords, rep2@coords, rep3@coords, rep4@coords) %>% magrittr::set_rownames(NULL))
sce$sample <- rep(paste0("mouse_", 1:4), times = c(ncol(rep1@counts), ncol(rep2@counts), ncol(rep3@counts), ncol(rep4@counts)))
logcounts(sce) <- transformGamPoi::shifted_log_transform(sce)
sce$plaque_density <- c(plaque_dens_rep1, plaque_dens_rep2, plaque_dens_rep3, plaque_dens_rep4)

# Remove lowly expressed genes
sce <- sce[rowSums(counts(sce)) > 50, ]
sce
```


Add gene ids
```{r, paged.print = FALSE}
eg <- clusterProfiler::bitr(rownames(sce), fromType = "SYMBOL", toType = "ENSEMBL", OrgDb="org.Mm.eg.db") %>%
  summarize(ENSEMBL = dplyr::first(ENSEMBL), .by = SYMBOL)
  
rowData(sce) <- data.frame(symbol = rownames(sce)) %>%
  tidylog::left_join(dplyr::rename(eg, symbol = SYMBOL, gene_id = ENSEMBL)) 
```

```{r}
cont2ordered <- function(x, n, from = min(x), to = max(x), add_label = TRUE, label_fmt = "%.2f"){
  groups <- seq(from, to, length.out = n+1)
  res <- rep(NA_integer_, length(x))
  labels <- rep(NA_character_, n)
  for(idx in seq_len(n)){
    if(idx == n){
      labels[idx] <- paste0("[", sprintf(label_fmt, groups[idx]), ",", sprintf(label_fmt, groups[idx+1]), "]")
      res[x >= groups[idx] & x <= groups[idx+1]] <- idx
    }else{
      labels[idx] <- paste0("[", sprintf(label_fmt, groups[idx]), ",", sprintf(label_fmt, groups[idx+1]), ")")
      res[x >= groups[idx] & x < groups[idx+1]] <- idx
    }
  }
  if(add_label){
    ordered(res, levels = seq_len(n), labels = labels)
  }else{
    res
  }
}

cont2ordered(seq(0, 30), n = 6, label_fmt = "%.f")
levels(cont2ordered(sce$plaque_density, n = 10, label_fmt = "%.2f"))

```



Add cell type labels to each cell

```{r}
brain_atlas_data_file <- "../data/cable_spatial_plaque_data/metacells.BrainCellAtlas_Saunders_version_2018.04.01.RDS"
brain_atlas_metadata_file <- "../data/cable_spatial_plaque_data//annotation.BrainCellAtlas_Saunders_version_2018.04.01.RDS"
if(! file.exists(brain_atlas_data_file) || ! file.exists(brain_atlas_metadata_file)){
  # See http://dropviz.org/ for more information
  data_url <- "https://storage.googleapis.com/dropviz-downloads/static/metacells.BrainCellAtlas_Saunders_version_2018.04.01.RDS"
  annotation_url <- "https://storage.googleapis.com/dropviz-downloads/static/annotation.BrainCellAtlas_Saunders_version_2018.04.01.RDS"
  
  download.file(data_url, brain_atlas_data_file)
  download.file(annotation_url, brain_atlas_metadata_file)
}
metacells_saunders <- readRDS(brain_atlas_data_file)
metacells_annotation <- readRDS(brain_atlas_metadata_file)
brain_ref <- SingleCellExperiment(list(counts = metacells_saunders), colData = metacells_annotation)
logcounts(brain_ref) <- transformGamPoi::shifted_log_transform(counts(brain_ref))
hippocampus_ref <- brain_ref[, brain_ref$tissue == "HC"]

length(intersect(rownames(brain_ref), rownames(sce)))
length(rownames(brain_ref))
length(setdiff(rownames(sce), rownames(brain_ref)))
# str(setdiff(rownames(sce), rownames(brain_ref)))

pred <- SingleR::SingleR(sce, ref = hippocampus_ref, labels = hippocampus_ref$class)
sce$CELL_TYPE <- pred$labels
sce$CELL_TYPE_LUMPED <- fct_lump(sce$CELL_TYPE, n = 5)

SingleR::plotScoreHeatmap(pred)
```

# Fine map the Hippocampus cell types

Download files

```{r}
yao_file <- "../data/cable_spatial_plaque_data/yao_isocortex_and_hippocampus_atlas.h5ad"
if(! file.exists(yao_file)){
  options(timeout = 3000)
  yao_dataset_url <- "https://datasets.cellxgene.cziscience.com/efbc53b2-559e-42cb-804d-f67443034cde.h5ad"
  download.file(yao_dataset_url, yao_file)
}
```

Subset the file (unfortunately, this has to happen in python)

```{r}
yao_sub_file <- "../data/cable_spatial_plaque_data/yao_hippocampus_atlas.h5ad"
if(! file.exists(yao_sub_file)){
  Sys.setenv(RETICULATE_PYTHON = "/g/huber/users/ahlmanne/conda-envs/pylemur_env/bin/python")
  Sys.setenv(PYTHONPATH = "/g/huber/users/ahlmanne/conda-envs/pylemur_env/bin/python")
  reticulate::use_condaenv("pylemur_env", required = TRUE)
  ad <- reticulate::import("anndata")
  adata <- ad$read_h5ad(yao_file)
  sub_adata <- adata[adata$obs["tissue"] == "hippocampal formation",]
  sub_adata$write_h5ad(yao_sub_file)
}

hippo_ref <- zellkonverter::readH5AD(yao_sub_file)
counts(hippo_ref) <- assay(hippo_ref, "X")
assay(hippo_ref, "X") <- NULL
logcounts(hippo_ref) <- transformGamPoi::shifted_log_transform(counts(hippo_ref))

match_rownames <- as_tibble(rowData(hippo_ref), rownames = "gene_id") %>%
  tidylog::left_join(as_tibble(rowData(sce)), by = "gene_id") %>%
  drop_na(symbol) %>%
  group_by(symbol) %>%
  summarize(gene_id = first(gene_id)) %>%
  dplyr::select(gene_id, symbol) %>%
  deframe()


length(setdiff(names(match_rownames), rownames(hippo_ref)))
hippo_ref <- hippo_ref[rownames(hippo_ref) %in% names(match_rownames), ]
rownames(hippo_ref) <- match_rownames[rownames(hippo_ref)]
```

```{r}
convert_interval_to_number <- function(interval){
  str_mat <- str_match(interval, ".(\\d+\\.?\\d*),(\\d+\\.?\\d*)")[,2:3]
  num_mat <- array(as.numeric(str_mat), dim = dim(str_mat))
  rowMeans(num_mat)
}

hvg_a <- order(-rowVars(logcounts(hippo_ref)))

a_sce <- hippo_ref[hvg_a[1:500], ]
b_sce <- sce[rownames(a_sce),]
b_sce$plaque_cluster <- cont2ordered(b_sce$plaque_density, n = 10, label_fmt = "%.2f")
b_sce$plaque_cluster_cont <- convert_interval_to_number(b_sce$plaque_cluster)
logcounts(a_sce) <- NULL
logcounts(b_sce) <- NULL
colData(a_sce) <- DataFrame(cell_id = colnames(a_sce),
                            origin = rep("ref", times = ncol(a_sce)), 
                            cell_type = colData(hippo_ref)$cell_type,
                            subclass = colData(hippo_ref)$subclass_label,
                            class = colData(hippo_ref)$class_label,
                            plaque_cluster_cont = 0)
colData(b_sce) <- DataFrame(cell_id = colnames(b_sce),
                            origin = rep("sce", times = ncol(b_sce)), 
                            cell_type = rep(NA, times  = ncol(b_sce)),
                            subclass = rep(NA, times  = ncol(b_sce)),
                            class = rep(NA, times  = ncol(b_sce)),
                            plaque_cluster_cont = b_sce$plaque_cluster_cont)
rowData(a_sce) <- NULL
rowData(b_sce) <- NULL
reducedDims(a_sce) <- NULL
reducedDims(b_sce) <- NULL
joint_sce <- cbind(a_sce, b_sce)
logcounts(joint_sce) <- transformGamPoi::shifted_log_transform(joint_sce)

set.seed(1)
integration_fit <- lemur::lemur(joint_sce, design = ~ origin + plaque_cluster_cont, n_embedding = 30, test_fraction = 0, verbose = TRUE)
integration_fit <- lemur::align_harmony(integration_fit)

new_df <- lemur.utils::transfer_col_data(integration_fit[,integration_fit$colData$origin == "ref"],
                               integration_fit[,integration_fit$colData$origin == "sce"], 
                               columns = vars(cell_type, subclass, class), k = 20) %>%
  mutate(cell_id = integration_fit$colData$cell_id[integration_fit$colData$origin == "sce"])

colData(sce) <- colData(sce) %>%
  as_tibble(rownames = "cell_id") %>%
  left_join(new_df, by = "cell_id") %>%
  DataFrame()
  
```

```{r}
table(sce$CELL_TYPE, sce$cell_type)
```


```{r}
int_umap <- uwot::umap(t(integration_fit$embedding))
as_tibble(colData(integration_fit)) %>%
  mutate(cell_type = fct_lump(cell_type,  n = 6)) %>%
  mutate(subclass = fct_lump(subclass,  n = 6)) %>%
  mutate(umap = int_umap) %>%
  sample_frac(n = 1) %>%
  ggplot(aes(x = umap[,1], y = umap[,2])) +
    geom_point(aes(color = subclass), size = 0.01) +
    facet_wrap(vars(origin)) +
    guides(color = guide_legend(override.aes = list(size = 2)))
```




```{r}
hist(sce$plaque_density)
```


```{r, paged.print=FALSE}
qc_df <- scuttle::perCellQCMetrics(sce, subsets = list(Mito = str_starts(rowData(sce)$symbol, "mt-")))
hist(qc_df$subsets_Mito_percent, breaks = 100)
hist(qc_df$sum, breaks = 100)

sce_subset <- sce[, qc_df$sum > 200 & qc_df$sum <= 5000 & qc_df$subsets_Mito_percent < 20]
sce_subset
```



Calculate UMAP
```{r, paged.print=FALSE}
set.seed(1)
umap <- scater::calculateUMAP(sce_subset)
reducedDim(sce_subset, withDimnames = FALSE) <- umap
```


```{r}
sce_subset$plaque_cluster <- cont2ordered(sce_subset$plaque_density, n = 10, label_fmt = "%.2f")
```




```{r}
as_tibble(colData(sce_subset)) %>%
  mutate(umap = umap) %>%
  ggplot(aes(x = umap[,1], y = umap[,2])) +
    geom_point(aes(color = sample), size = 0.3, stroke = 0) +
    guides(color = guide_legend(override.aes = list(size = 1)))

as_tibble(colData(sce_subset)) %>%
  mutate(umap = umap) %>%
  ggplot(aes(x = umap[,1], y = umap[,2])) +
    geom_point(aes(color = CELL_TYPE_LUMPED), size = 0.3, stroke = 0) +
    guides(color = guide_legend(override.aes = list(size = 1)))

as_tibble(colData(sce_subset)) %>%
  mutate(umap = umap) %>%
  ggplot(aes(x = umap[,1], y = umap[,2])) +
    geom_point(aes(color = plaque_density), size = 0.3, stroke = 0) +
    scale_color_viridis_b()
```


```{r}
set.seed(1)
fit <- lemur::lemur(sce_subset, design = ~ plaque_cluster, n_embedding = 30, test_fraction = 0.6)
set.seed(1)
fit <- lemur::align_harmony(fit)
```


```{r}
fit <- lemur::test_de(fit, cond(plaque_cluster = "[0.90,1.00]") - cond(plaque_cluster = "[0.00,0.10)"))
set.seed(1)
nei <- lemur::find_de_neighborhoods(fit, group_by = vars(plaque_cluster, sample), test_method = "edgeR")
```


```{r}
qs::qsave(sce_subset, "../data/cable_spatial_plaque_data/annotated_sce.qs")
```

```{r}
fit_small <- fit["Jun",]
nei_small <- nei %>% filter(name == "Jun")

qs::qsave(fit_small, "../output/cable_spatial_plaque_data/fit_small.qs")
qs::qsave(nei_small, "../output/cable_spatial_plaque_data/nei_small.qs")
```


# Session Info

```{r}
sessionInfo()
```


